---
title: "Kronecker method for additive GP regression example with stan"
output: html_notebook
---

This is an R notebook demonstrating the Kronecekr method for additive Gaussian process model. The model is implemented in stan. Here we use `GP_3d_kron_alltwoway_est.stan` with `cmdstanr` package - check out [this page](https://mc-stan.org/cmdstanr/) to install `cmdstan`

```{r}
path = '~/GitHub/Additive-GP-Kronecker/'
path_stan = '~/GitHub/Additive-GP-Kronecker/Code/Stan/'
library(ggplot2)
library(plyr)
library(cmdstanr)
```

Load datasets. We need
* NO2_imputation_STadjust.csv - data after manipulation (missing value imputation, adjustment of winter->summer time)
* London_air_station.csv - Information about the monitoring stations (longitude, latitude, etc)

```{r}
df_NO2 = read.csv(paste0(path,'/Data/NO2_imputation_STadjust.csv'), header=TRUE)
df_site = read.csv(paste0(path,'/Data/London_air_station.csv'), header=TRUE)
y_wide = df_NO2[,2:60] 
df_site = df_site[df_site$SiteCode%in%colnames(y_wide),]
df_site$SiteCode == colnames(y_wide)
```

### Hyper-parameter estimation by MCMC
Extraxt necessary data to feed to stan file
```{r}
S = as.matrix(cbind(df_site$Easting,df_site$Northing))/10000
N1 = dim(S)[1]
N2= dim(y_wide)[1]/24
N3 = 24
Dy = as.matrix(1:N2,1,N2)
Hr = as.matrix(0:23,1,24)
D1 = dim(S)[2]
D2 = dim(Dy)[2]
D3 = dim(Hr)[2]
N = N1*N2*N3
y = c(as.matrix(y_wide))
Hurst = c(0.3,0.5,0.5)
```
Read and compile, compiling may takes some time! 
```{r}
file = paste0(path_stan, 'GP_3d_kron_alltwoway_est.stan')
mod = cmdstan_model(file,  include_paths = path_stan)
```
Sampling
```{r}
data = list(N1= N1, N2=N2, N3=N3,
            NN1= as.integer(N/N1), NN2=as.integer(N/N2), NN3=as.integer(N/N3), 
            D1=D1, D2=D2, D3=D3, 
            X1 = S, X2 = Dy, X3 = Hr,  y=y, 
            Hurst1=Hurst[1],Hurst2=Hurst[2], Hurst3=Hurst[3])
fit = mod$sample(
  data = data, 
  seed = 281, 
  iter_warmup = 200,
  iter_sampling = 300,
  save_warmup = TRUE,
  chains = 2, 
  parallel_chains = 2,
  refresh = 10
)
```

Saving the output to csv if needed (make sure you have Output folder!)
```{r}
fit$save_output_files(dir = paste0(path,"/Output"), basename = NULL, timestamp = TRUE, random = TRUE)
```

Summary, traceplot etc.
```{r}
fit$summary(variables = c('alpha0','alpha1','alpha2','alpha3','sigma','lp__'))
post_mean = fit$summary(variables = c("alpha0",'alpha1','alpha2','alpha3','sigma'))$mean
```

```{r}
post_samples = fit$draws(format = "df",  inc_warmup = FALSE)
post_samples$.chain = as.character(post_samples$.chain)
gg = ggplot(data=post_samples, aes(x=.iteration, y = alpha3, color=.chain))+
  geom_line() 
gg
```

### Prediction

```{r}
post_mean = fit$summary(variables = c("alpha0",'alpha1','alpha2','alpha3','sigma'))$mean
```


