---
title: "Kronecker method for additive GP regression example with stan"
output: html_notebook
---

This is an R notebook demonstrating the Kronecekr method for additive Gaussian process model. The model is implemented in stan. Here we use `GP_3d_kron_alltwoway_est.stan` with `cmdstanr` package - check out [this page](https://mc-stan.org/cmdstanr/) to install `cmdstan`

```{r}
path = '~/GitHub/Additive-GP-Kronecker/'
path_stan = '~/GitHub/Additive-GP-Kronecker/Code/Stan/'
library(ggplot2)
library(plyr)
library(cmdstanr)
library(rstan)
```

Load datasets. We need
* NO2_imputation_STadjust.csv - data after manipulation (missing value imputation, adjustment of winter->summer time)
* NO2_STadjust.csv - data after manipulation (no missing value imputation but with adjustment of winter->summer time)
* London_air_station.csv - Information about the monitoring stations (longitude, latitude, etc)
`NO2_STadjust.csv` is only needed for the prediction / posterior procedure

```{r}
df_NO2 = read.csv(paste0(path,'/Data/NO2_imputation_STadjust.csv'), header=TRUE)
df_NO2_observed = read.csv(paste0(path,'/Data/NO2_STadjust.csv'), header=TRUE)
df_site = read.csv(paste0(path,'/Data/London_air_station.csv'), header=TRUE)
y_wide = df_NO2[,2:60] 
df_site = df_site[df_site$SiteCode%in%colnames(y_wide),]
df_site$SiteCode == colnames(y_wide)
```

### Hyper-parameter estimation by MCMC
Extract necessary data to feed to stan file
```{r}
S = as.matrix(cbind(df_site$Easting,df_site$Northing))/10000
N1 = dim(S)[1]
N2= dim(y_wide)[1]/24
N3 = 24
Dy = as.matrix(1:N2,1,N2)
Hr = as.matrix(0:23,1,24)
D1 = dim(S)[2]
D2 = dim(Dy)[2]
D3 = dim(Hr)[2]
N = N1*N2*N3
y = c(as.matrix(y_wide))
Hurst = c(0.3,0.5,0.5)
```
Read and compile, compiling may takes some time.
```{r}
file = paste0(path_stan, 'GP_3d_kron_alltwoway_est.stan')
mod = cmdstan_model(file,  include_paths = path_stan)
```
Sampling
```{r}
data = list(N1= N1, N2=N2, N3=N3,
            NN1= as.integer(N/N1), NN2=as.integer(N/N2), NN3=as.integer(N/N3), 
            D1=D1, D2=D2, D3=D3, 
            X1 = S, X2 = Dy, X3 = Hr,  y=y, 
            Hurst1=Hurst[1],Hurst2=Hurst[2], Hurst3=Hurst[3])
fit = mod$sample(
  data = data, 
  seed = 281, 
  iter_warmup = 200,
  iter_sampling = 300,
  save_warmup = TRUE,
  chains = 2, 
  parallel_chains = 2,
  refresh = 10
)
```

Saving the output to csv if needed (make sure you have Output folder)
```{r}
fit$save_output_files(dir = paste0(path,"/Output"), basename = NULL, timestamp = TRUE, random = TRUE)
```

Summary, traceplot etc.
```{r}
fit$summary(variables = c('alpha0','alpha1','alpha2','alpha3','sigma','lp__'))
post_mean = fit$summary(variables = c("alpha0",'alpha1','alpha2','alpha3','sigma'))$mean
```

```{r}
post_samples = fit$draws(format = "df",  inc_warmup = FALSE)
post_samples$chain = as.character(post_samples$.chain)
gg = ggplot(data=post_samples, aes(x=.iteration, y = alpha3, color=chain))+
  geom_line() + theme_minimal() 
gg
```

### Prediction / Posterior

```{r}
post_mean = fit$summary(variables = c("alpha0",'alpha1','alpha2','alpha3','sigma'))$mean
post_mean
```
Test data - here we look into 4 monitoring stations picked in the paper
```{r}
site_list_test = c('CR5','TH4','KC1','BT5') # select monitoring stations of interest
y_wide_test = df_NO2_observed[,site_list_test]
df_site_test = df_site[df_site$SiteCode %in% site_list_test,]
# make sure that data is in the right order
df_site_test = df_site_test[match(site_list_test, df_site_test$SiteCode),]
df_site_test$SiteCode ==colnames(y_wide_test)
S_test = as.matrix(cbind(df_site_test$Easting,df_site_test$Northing))/10000
N1_test = dim(S_test)[1]
N2_test = dim(y_wide)[1]/24 # if you are interested in shorter period, you can change here!
N3_test = 24
N_test = N1_test*N2_test*N3_test
Dy_test = as.matrix(1:N2_test,1,N2_test)
Hr_test = as.matrix(0:23,1,24)
y_test = c(as.matrix(y_wide_test))
```
Posterior (predictive) mean and variance. We do not have to use Stan for this unless you are interested in sampling from posterior, but since we have all the codes and helper functions written in Stan from estimation, let's stick with Stan for now. A code that is written solely in R will be uploaded soon.
```{r}
file = paste0(path_stan,'/GP_3d_kron_alltwoway_pred.stan')
mod = cmdstan_model(file)
data = list(N1= N1, N2=N2, N3=N3, 
            NN1= as.integer(N/N1), NN2=as.integer(N/N2), NN3=as.integer(N/N3), 
            D1=D1, D2=D2, D3=D3,
            X1 = S, X2 = Dy, X3=Hr, y=y, 
            Hurst1=Hurst[1],Hurst2=Hurst[2], Hurst3=Hurst[3],
            N1_new= N1_test, N2_new=N2_test, N3_new=N3_test, 
            X1_new = S_test,  X2_new = Dy_test, X3_new=Hr_test,
            alpha0 = post_mean[1], alpha1 = post_mean[2], alpha2 = post_mean[3], 
            alpha3 = post_mean[4],  sigma = post_mean[5]
            )
pred = mod$sample(
  data = data, 
  iter_warmup = 0,
  iter_sampling = 1,
  chains = 1, 
  refresh = 1,
  fixed_param = TRUE
)
pred$save_output_files(dir = paste0(path0,"/Output"), basename = NULL, timestamp = TRUE, random = TRUE)
predfit = rstan::read_stan_csv(pred$output_files()) # changing to stanfit object 
```
Now we can extract the component-wise mean & variance. To get overall fitted value, we can simply add component-wise means but this doesn't work with variance, as we have to take into account the covariance between l-th term and l'-th term. If overall variance is of interest, you can modify the Stan code, see `GP_3d_kron_alltwoway_pred_variance.stan` file for example.
```{r}
n_terms = 7 # change here depending on the model, we have n_term = 8 for saturated model
cnames = c('const','space','day','hour', 'space_day','space_hour', 'day_hour','space_day_hour')
params = c('mu0','mu1','mu2','mu3','mu12','mu13','mu23','mu123')
f_mean = as.data.frame(matrix(NA, N_test,n_terms))
colnames(f_mean) = cnames[1:n_terms]
#main terms
f_mean$const = rep(colMeans(as.matrix(predfit, pars = c(params[1]))), times=N_test)
f_mean$space = rep(colMeans(as.matrix(predfit, pars = c(params[2]))), each=(N2_test*N3_test))
f_mean$day = rep(rep(colMeans(as.matrix(predfit, pars = c(params[3]))),each = N3_test),times=N1_test)
f_mean$hour = rep(colMeans(as.matrix(predfit, pars = c(params[4]))), times=(N1_test*N2_test))
# two way
f_mean$space_day = rep(colMeans(as.matrix(predfit, pars = c(params[5]))), each=N3_test)
f_mean$space_hour = rep(0, N_test)
  t13tmp = colMeans(as.matrix(predfit, pars = c(params[6])))
  for (n1 in 1:N1_test){
    N23 = N3_test*N2_test
    f_mean$space_hour[((n1-1)*N23+1):(n1*N23)] = rep(t13tmp[(((n1-1)*N3_test)+1):(n1*N3_test)],times=N2_test)
  }
  rm(t13tmp)
f_mean$day_hour = rep(colMeans(as.matrix(predfit, pars = c(params[7]))), times=N1_test)

## three way -- only needed when running saturated model
# fitted_mean$space_day_hour = colMeans(as.matrix(predfit, pars = c(params[8]))) 

# all
f_mean$all = rowSums(f_mean)
f_mean$observed = as.vector(as.matrix(y_test))
f_mean$residual = f_mean$observed - f_mean$all
f_mean$site = rep(site_list_test, each = N23)
f_mean$date = rep(df_NO2$date, N1_test)
```
For variance, repeat the above with `params = c('v0','v1','v2','v3','v12','v13','v23','v123')`. 

### Some plots
To show hour of the day effect by sites 
```{r}
df_hour = as.data.frame(cbind(rowSums(f_mean[,c('hour','space_hour')]), rep(0:23, length(site_list_test))))
colnames(df_hour) = c('NO2','hour')
df_hour$site = factor(f_mean$site,levels = site_list_test)
gg = ggplot(df_hour, aes(x=hour, y=NO2, color=site, linetype=site)) +
    geom_line() + ylim(-20,15) +
    labs(title = 'Hour of the day effect', x='Hour', y=NULL) +
    scale_x_continuous(breaks = seq(0,24, by = 4)) +
    theme_minimal(base_size = 14)
print(gg)
```
To show global time trend by sites
```{r}
## To highlight weekend
weekend_st_day = seq(as.POSIXct("2020/1/11", tz='GMT'), as.POSIXct("2020/5/31", tz='GMT'), by = '7 days')
weekend_en_day = seq(as.POSIXct("2020/1/13", tz='GMT'), as.POSIXct("2020/6/1", tz='GMT'), by = '7 days')
week_st = 1
week_en = 21
num_weeks = week_en - week_st + 1
rects = data.frame(xstart = weekend_st_day[week_st:week_en], xend=weekend_en_day[week_st:week_en], col = rep('Weekend',num_weeks))

for (s in site_list_test){
  df_plot_site = f_mean[f_mean$site==s,]
  df_day_bysite = as.data.frame( matrix(ncol = 2, nrow=(num_weeks*7)))
  colnames(df_day_bysite) = c('NO2','date')
  df_day_bysite$NO2 = rowSums(df_plot_site[seq((24*7*(week_st-1)+1),(24*7*week_en),by=24),c('day','space_day')])
  df_day_bysite$date = seq(as.POSIXct("2020/1/6", tz='GMT'), as.POSIXct("2020/5/31", tz='GMT'), by = 'days')
  gg = ggplot() +
    geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.3) +
    geom_line(data = df_day_bysite, aes(x=date, y=NO2), show.legend = FALSE) +
    ylim(-35, 65) + 
    labs(title = NULL, subtitle = paste('Site:', s), x=NULL, y=NULL) +
    guides(fill = FALSE, color = FALSE, linetype = FALSE, shape = FALSE) +
    scale_x_datetime(date_breaks="14 days", 
                     labels = scales::label_date("%b %d", tz="GMT"),
                     guide = guide_axis(angle = 45)) +
    theme_minimal(base_size = 14) 
  print(gg)  
}
```
